Concurrent and Distributed Programming in Harmony

### 2022-01-11 paper/paper.tex
Latest

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Inprecisely, methods that use the <{atomically}> keyword are specifications
while methods that do not are probably implementation.
Whether something is a specification or an implementation is in the
eye of the beholder.
Importantly, however, is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of implementation of
\autoref{fig:synch}(b) against the specification of
\autoref{fig:synch}(a).

@@

\paragraph{SetIntLevel.}  A context contains a boolean that determines
if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
new interrupt level and pushes the old.

@@

However, users first want to focus on the rows in the top table.
In this particular example, there are 6 threads, the thread that
first runs and the five threads that are spawned by it.  They take
``turns'' by context switching.  In this particular trace, <{diner(2)}>
runs first and acquires <{forks[2]}>, its left fork.  In the table
in the lower-right, you can see that the thread is currently blocked
and in atomic as well as read-only mode (conditions in a <{when}>
clause are not allowed to modify state).  You can also see its stack
trace: it trying to acquire <{forks[3]}>, its right fork (executing
the code in \autoref{fig:synch}(a)).  Next <{diner(0)}>, <{diner(3)}>,
<{diner(4)}>, and <{diner(1)}> take similar turns.  All threads are
blocked waiting for another thread to release a lock---a classic
deadlock situation.

@@

Combinatorial state explosion.
Production code.
Lack of formal definition of Harmony itself.

>>>>

For example, \autoref{fig:synch}(a) shows the specification for locks
from the Harmony <{synch}> module that was used in \autoref{fig:ms},
featuring the Harmony <{when}> statement.
\autoref{fig:synch}(b) shows an implementation of the <{acquire}>
method based on a test-and-set primitive, which in turn is
\emph{specified}.
Inprecisely, methods that use the <{atomically}> keyword are specifications
while methods that do not are probably implementation.
Whether something is a specification or an implementation is in the
eye of the beholder.
Important, however, is the notion of modularity that is implied here.
Harmony can model check the implementation of \autoref{fig:ms} using
the specification of a lock in \autoref{fig:synch}(a).
Separately, Harmony can check the correctness of implementation of
\autoref{fig:synch}(b) against the specification of
\autoref{fig:synch}(a).  This modularity allows larger systems to
be model checked that would not be tractable otherwise.

@@

\paragraph{SetIntLevel.}  A context contains a boolean that determines
if interrupts are enabled or not.  \texttt{SetIntLevel} pops the
new interrupt level and pushes the old.  Supporting interrupts is
important.  First, writing multi-threaded programs that also need
to handle interrupts is difficult.  Second, distributed algorithms
often require timer interrupts to trigger actions such as retransmissions.

@@

However, users first want to focus on the rows in the top table.
In this particular example, there are 6 threads, the thread that
first runs and the five threads that are spawned by it.  They take
``turns'' by context switching.  In this particular trace, <{diner(2)}>
runs first and acquires <{forks[2]}>, its left fork.  In the table
in the lower-right, you can see that the thread is currently blocked
and in atomic as well as read-only mode (conditions in a <{when}>
clause are not allowed to modify state).  You can also see its stack
trace: it is trying to acquire <{forks[3]}>, its right fork (executing
the code in \autoref{fig:synch}(a)).  Next <{diner(0)}>, <{diner(3)}>,
<{diner(4)}>, and <{diner(1)}> take similar turns.  All threads are
blocked waiting for another thread to release a lock---a classic
deadlock situation.

@@

While Harmony has proven itself to be effective and even fun to use
for a large variety of synchronization problems and distributed
algorithms, it is not a panacea.  At the core of Harmony is a model
checker and with all its optimizations it can still easily suffer
from combinatorial state explosion in programs that have too many
threads or too many non-deterministic choices.  Nonetheless, because
of its systematic state exploration, many problems can often be
uncovered in small scenarios with few threads and few choices.

The Harmony language, at least currently, is not usable for production
code.  One reason for this is the <{atomically}> clause.  On a multi-core
processor it would require an instruction to stop all other cores.
Harmony was never designed to be a language for production code.  Because
it is close to languages such as Python and C, it tends to be fairly
easy to hand-translate code developed in Harmony to a desired target
programming language.  However, such hand-translation could easily
introduce a bug that did not exist in the Harmony code.

Finally, the Harmony language currently does not have a formal definition
itself, and there is no guarantee that the compiler itself is correct.
Harmony compiles to a virtual machine code that is fully formally
specified in TLA+, and this is the level of abstraction where model
checking occurs.
